---
title: "eds223_hw4_Aoverbye"
author: "Amanda G. Overbye"
output: 
  html_document:
    code_folding: true
editor: visual
execute: 
  eval: true
  warning: false
editor_options: 
  chunk_output_type: console
---

## Homework Assignment 4

#### Prioritizing potential aquaculture

```{r}
# Load libraries
library(tidyverse)
library(here) 
library(tmap)
library(ggplot2)
library(stars)
library(spDataLarge)
```

# Prepare data

```{r}
# Load data

wc_regions_clean <- st_read(here("data", "Wc_regions_clean.shp"))
```

### Preliminary Data Exploration

```{r}
print(colnames(wc_regions_clean))
```

# Process data

# Find suitable locations

# Determine the most suitable EEZ

## Creating A Function

"To make your workflow generalizable, you must create a function that has the following characteristics:

**arguments:**

-   minimum and maximum sea surface temperature
-   minimum and maximum depth species name

**outputs:**

-   map of EEZ regions colored by amount of suitable area species
    -   species name should be included in the map’s title
