---
title: "eds223_hw4_Aoverbye"
author: "Amanda G. Overbye"
output: 
  html_document:
    code_folding: true
editor: visual
execute: 
  eval: true
  warning: false
editor_options: 
  chunk_output_type: console
---

## Homework Assignment 4

#### Prioritizing potential aquaculture

# Creating A Function

"To make your workflow generalizable, you must create a function that has the following characteristics:

**arguments:**

-   minimum and maximum sea surface temperature
-   minimum and maximum depth species name

**outputs:**

-   map of EEZ regions colored by amount of suitable area species
    -   species name should be included in the map’s title

```{r}
# Load libraries
library(tidyverse)
library(here) 
library(tmap)
library(ggplot2)
library(stars)
library(spDataLarge)
```

## Prepare data

```{r}
# Load data
# Read in NOAA Sea Surface Temperature files
sst_avg2008 <- read_stars(here("data", "average_annual_sst_2008.tif"))
sst_avg2009 <- read_stars(here("data", "average_annual_sst_2009.tif"))
sst_avg2010 <- read_stars(here("data", "average_annual_sst_2010.tif"))
sst_avg2011 <- read_stars(here("data", "average_annual_sst_2011.tif"))
sst_avg2012 <- read_stars(here("data", "average_annual_sst_2012.tif"))

# Read in Bathymetry files
depth <- read_stars(here("data", "depth.tif"))

# Load wc_regions shape file
wc_regions_clean <- st_read(here("data", "Wc_regions_clean.shp"))
```

### Preliminary Data Exploration

My first step in this project is getting an overview of the data I am. I do this by looking at the first 5 rows(head) of any shape files and by doing a quick plot of the tif files.

```{r}
# View the column names of the wc_regions data
print(colnames(wc_regions_clean))
print(head(wc_regions_clean))
```

```{r}
# Plot bathymetry data
plot(depth)
```

I am only going to show one plot of the surface temperature data because that is all I need right now to get an idea of how it will look.

```{r}
# Plot the temperature data
plot(sst_avg2008)
```

#### CRS systems

One of things I have learned is that mismatching CRSs is one of the things that will derail me the most when I am trying to work with geographical data sets. Lots of weird things can happen when coding and I think in projects like these about 90% of those things are caused by CRS troubles. If I were ever caught in the bad place, one of the signs would be mis-matched CRSs.

In order to prevent these troubles, I will write a function that will tell me if data sets have mismatching CRSs

```{r}
# Function to determine if CRSs match
check_crs_fun <- function(d1, d2) {
  if (identical(st_crs(d1), st_crs(d2))) {
    return(print("CRSs match!"))
  } else {
    return(print("CRSs do not match"))
  }
}
```

Then we can check our data sets:

```{r}
check_crs_fun(depth, sst_avg2008)
check_crs_fun(depth, wc_regions_clean)
check_crs_fun(wc_regions_clean, sst_avg2008)
check_crs_fun(sst_avg2009, sst_avg2008)
```

Here we can see that none of our CRSs match.

Our next step will be to change the CRSs so they are consistent. We will do this by creating a function that will change the CRS of the first inputted dataset to match the crs of the second inputted dataset. I am going to change the CRSs of the `depth` and `wc_regions_clean` datasets to match the `sst_avg` datasets.

```{r}
change_crs_fun <- function(df1, df2) {
  if (identical(st_crs(df1), st_crs(df2))) {
    print("no change")
    return(df1)
  } else {
    df1 <- st_transform(df1, st_crs(df2))
    print("changed first dataset's CRS to match the second dataset's CRS")
    return(df1)
  }
}
```

```{r}
# Change the CRSs to match and store in new varibles
changed_depth <- change_crs_fun(depth, sst_avg2008)
changed_wc_regions_clean <- change_crs_fun(depth, sst_avg2008)
```

```{r}
# Check if the CRSs match now 
check_crs_fun(changed_depth, sst_avg2008)
check_crs_fun(wc_regions_clean, sst_avg2008)
```

#### Stacking Rasters

# **Write about why stacking rasters is useful**
# ** Write about why we crop rasters**
```{r}
# Crop all rasters to the extent of the first raster
sst_avg2009 <- st_crop(sst_avg2009, st_bbox(sst_avg2008))
sst_avg2010 <- st_crop(sst_avg2010, st_bbox(sst_avg2008))
sst_avg2011 <- st_crop(sst_avg2011, st_bbox(sst_avg2008))
sst_avg2012 <- st_crop(sst_avg2012, st_bbox(sst_avg2008))
```

```{r}
# Stacking Rasters

sst_avg_stack <- c(sst_avg2008, sst_avg2009, sst_avg2010, sst_avg2011, sst_avg2012)
```


Now that we know we have matching CRSs and our rasters are stacked we can move to the next step..

## Processing The data!

## Find suitable locations

## Determine the most suitable EEZ

## Citations

#### Data

Flanders Marine Institute (VLIZ). (n.d.). *Exclusive Economic Zone (EEZ) boundaries*. Maritime Boundaries Database, version unknown. Retrieved from https://www.marineregions.org/eez.php (Accessed: November 29, 2024)

General Bathymetric Chart of the Oceans (GEBCO) Compilation Group. (2022). *GEBCO_2022 Grid* \[Data set\]. https://doi.org/10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c (Accessed: November 29, 2024)

National Oceanic and Atmospheric Administration (NOAA). (2018). *Daily Global 5km Satellite Sea Surface Temperature Anomaly (Version 3.1)*. NOAA Coral Reef Watch. Retrieved from https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php (Accessed: November 29, 2024)
